Concurrent Programming with CXL

### 2020-06-04 book/paper.tex
Latest

This book is intended to help people with understanding and
developing concurrent code.  In particular, it uses a new tool
called CXL that help with \emph{testing} concurrent code.
The approach is based on \emph{model checking}: instead of relying
on luck, CXL will run \emph{all possible executions} of a particular
test program.  So even if a bug is unlikely to occur, if the test
\emph{can} expose the bug it \emph{will}.  Importantly, if the bug is
found, the model checker precisely shows how to trigger it in
the smallest number of steps.

@@

So why is this useful?  Consider, for example, a sorting algorithm.
Now suppose we create a test program, a model, that tries sorting
\emph{all} lists of up to five numbers chosen from the set \{ 1,
2, 3, 4, 5 \}.  Model checking proves that for those particular
scenarios the sorting algorithm works: the output is a sorted
permutation of the input.  In some sense it is an excellent test:
it will have considered all corner cases, including lists where all
numbers are the same, lists that are already sorted or reversely
sorted, etc.  If there is a bug in the sorting algorithm, most
likely it would be triggered and the model checker would produce a
scenario that would make it easy to find the source of the bug.
However, if the model checker does not find any bugs, we do not
know necessarily that the algorithm works for lists with more than
five numbers or for lists that have values other than the numbers
1 through 5.  Still, we would expect that the likelihood that there
are bugs remaining in the sorting algorithm is small.

But model checking can also help with proving an algorithm correct.
The reason is that the correctness of many programs is built around
\emph{invariants}: predicates that must hold for every state in the
execution of a program.  A model checker can find violations of
those invariants when evaluating a model and provide valuable early
feedback to somebody who is trying to construct a proof.
Throughout this book we will include, as much as possible, examples
of such invariants as they often provide excellent insight into
why a particular algorithm works.

So what is CXL?
CXL is a concurrent programming language.  It was designed to teach
the basics of concurrent programming, but it is also useful for
testing new concurrent algorithms or even sequential and distributed
algorithms.  CXL programs are not intended to be ``run'' like programs
in most other programming languages---instead CXL programs are
model checked to verify that the program has certain desirable
properties and does not suffer from bugs.

@@

CXL programs manipulates CXL values.
CXL values are recursively defined:
they include booleans (\texttt{False} and \texttt{True}),
integers (but not floating point numbers),
strings (enclosed by double quotes),
sets of CXL values,
and dictionaries that map CXL values to other CXL values.

Another type of CXL value is the \emph{atom}.  It is essentially
just a name.  An atom is denoted using a period followed by the
name.  For example, \texttt{.main} is an atom.

CXL makes extensive use of dictionaries.  Their syntax and properties
are a little different from Python.
A dictionary maps CXL values, known as \emph{keys}, to CXL values.
Unlike Python, any CXL value can be a key, including another
dictionary.
CXL dictionaries are written as
$\mathtt{dict}\{ k_0: v_0, ~ k_1: v_1, ~ ... \}$.
If \texttt{d} is a dictionary, and \texttt{k} is a key, then the
following expression retrieves the CXL value that $k$ maps to:
\begin{code}
\begin{verbatim}
d k
\end{verbatim}
\end{code}
This is unfamiliar, but in CXL square brackets can be used in the
same way as parentheses, so you can express it in the more familiar
form:
\begin{code}
\begin{verbatim}
d[k]
\end{verbatim}
\end{code}
However, if \texttt{d = dict\{ .count: 3 \}}, then you can write
can write \texttt{d.count} (which has value 3) instead of having to write
\texttt{d[.count]} (although both will work).
The meaning of \texttt{d a b c ...} is \texttt{((((d a) b) c) ...)}.

Tuples are special forms of dictionaries where the keys are
the indexes into the tuple.  For example, the tuple
\texttt{(5, False)} is the same CXL value as
\texttt{dict\{ 0: 5, 1: False \}}.
The empty tuple \texttt{()} is the same value as \texttt{dict\{\}}.
Note that this is different from the empty set, which is \texttt{\{\}}.
As in Python, you can create singleton tuples by adding an
extra ",". For example, \texttt{(1,)}.

@@

The state of a process is called a \emph{context}: it contains the value of
its name tag, program counter, stack, and register.
The state of the CXL machine
consists of the value of its memory and the multiset (or \emph{bag}) of
contexts.  It is a multiset of contexts because two different processes can
be in the same state.
The initial state of the CXL memory is the empty tuple, \texttt{()}.
The context bag has an initial context in it with name tag
\texttt{\_\_main\_\_/()}, pc 0, register \texttt{()}, and an empty stack.
Each machine instruction updates the state in some way.
Figure~\ref{fig:incstate} shows an example of a reachable state for 
the program in Figure~\ref{fig:inc}.

It may seem strange that there is only one memory location and that each
process has only one register.  However, this is not a limitation because
CXL values are unbounded trees.
Both the memory and the register of a process always contain
a special type of CXL value in which the root is a dictionary
that maps atoms to CXL values.  We call this a \emph{directory}.
A directory represents the state of a collection of variables named by the atoms.

Because directories are CXL values themselves,
directories can be organized into a tree.
Each node in the tree is then identified
by a sequence of atoms, like a path name in the file system hierarchy.  We call
such a sequence the \emph{address} of a CXL value.

Compiling the code in Figure~\ref{fig:inc} results in the CXL machine code
listed in Figure~\ref{fig:inccode}.
The CXL machine is predominantly a \emph{stack machine}.
Most instructions pop values from the stack or push values onto the stack.
Initially there is one process that starts executing at instruction 0.
In this case, there is a \texttt{JUMP} instruction that sets the program
counter to 22.  At program counter 1 is the code for the \texttt{up} method.
All methods start with a \texttt{Frame} instruction and end with a \texttt{Return}
instruction.

@@

\begin{enumerate} \setcounter{enumi}{1}
\item the \texttt{Push} instruction pushes the constant 1
onto the stack of the process.
\item the \texttt{PushAddress} instruction pushes the address of
the shared variable \texttt{count} onto the stack.
\item The \texttt{Load} instruction pops the address of the
\texttt{count} variable and then pushes the value of the
\texttt{count} variable onto the stack.
\item \texttt{2-ary} is a \texttt{+} operation with 2 arguments.
It pops two values from the stack (1 and the value of \texttt{count}),
adds them, and pushes the result back onto the stack.
\item The \texttt{PushAddress} instruction pushes the address of the \texttt{count}
variable onto the stack.
\item The \texttt{Store} instruction pops the address of a variable and pops
a CXL value (the sum of the \texttt{count} variable and 1), and updates the
\texttt{count} variable.
\end{enumerate}

Figure~\ref{fig:incoutput} shows the output produced by running the
\texttt{Up.cxl} program.
It starts by reporting that the assertion on line 9 in column 5 failed and that
\texttt{count} has value 1 instead of 0:
\texttt{Assertion failed (’assert’, ’Up.cxl’, 9, 5) 1}.
Next it reports an execution that failed this assertion.  The output has
four columns:
\begin{enumerate}
\item The name tag of the process;
\item The sequence of program counters of the CXL machine instructions that the process executed;
\item The current program counter of the process;
\item The contents of the shared memory.
\end{enumerate}

If we look at the middle three rows, we see that:
\begin{enumerate}
\item Process 0 executed instructions 1 through 7, loading the value of
\texttt{count} but stopping just before storing 1 into \texttt{count};
\item Process 1 executed instructions 1 through 12, storing 1 into
into \texttt{count} and storing \texttt{True} into \texttt{done[2]};
\item Process 0 continues execution, storing value 1 into \texttt{count}
and storing \texttt{True} into \texttt{done[1]}.
\end{enumerate}

@@

Hopefully you have started thinking of how to solve the concurrency
problem and you may already have prototyped some solutions.
In this chapter we will go through a few reasonable but broken attempts.
At the heart of the problem is that we would like make sure that when
the \texttt{count} variable is being updated that no other process is
trying to do the same thing.  We call this a \emph{critical section}: a
set of instructions where only one process is allowed to execute at a
time.

Critical sections are useful when accessing a shared data
structure, particularly when that access requires multiple underlying
machine instructions.  A simple counter is a very simple example of
a data structure.  A more involved one would be access to a binary tree.
Adding a node to a binary tree or re-balancing a tree often requires
multiple operations.  Maintaining consistency is certainly much easier
(although not necessarily impossible) if during this time no other
process also tries to access the binary tree.

A critical section is often modeled as processes in an infinite loop
entering and exiting the critical section.  In CXL:

\begin{code}
\begin{verbatim}
while True:
    @ncs: pass; # in non-critical section
    # enter critical section
    @cs: pass;  # in critical section
    # exit critical section
;
\end{verbatim}
\end{code}

Here \texttt{@ncs} and \texttt{@cs} are \emph{labels}, each identifying
a location in the CXL machine code.  The first thing we need to
ensure is that there cannot be two processes at the critical section.
We would like to place an assertion at the \texttt{@cs} label that
specifies that only the current process can be there.  CXL in fact
supports this.  It has an operator \texttt{atLabel $L$}, where $L$
is the atom of the label (in this case, \texttt{.cs}).  The
operator returns a bag of name tags of processes executing at that
label.  The bag is represented by a dictionary that maps each element
in the bag to the number of times the element appears in the bag.
Method \texttt{atLabel} only exists for specification purposes---do not
use it in normal code.

Figure~\ref{fig:csbarebones} shows the code.
The assertion also makes use of the \texttt{nametag()} operator that
returns the name tag of the current process.
If you run the code through CXL, the assertion should fail because
there is no code yet for entering and exiting the critical section.

>>>>

This book is intended to help people with understanding and
developing concurrent code.  In particular, it uses a new tool
called CXL that helps with \emph{testing} concurrent code.
The approach is based on \emph{model checking}: instead of relying
on luck, CXL will run \emph{all possible executions} of a particular
test program.  So even if a bug is unlikely to occur, if the test
\emph{can} expose the bug it \emph{will}.  Importantly, if the bug is
found, the model checker precisely shows how to trigger it in
the smallest number of steps.

@@

So why is this useful?  Consider, for example, a sorting algorithm.
Now suppose we create a test program, a model, that tries sorting
\emph{all} lists of up to five numbers chosen from the set \{ 1,
2, 3, 4, 5 \}.  Model checking proves that for those particular
scenarios the sorting algorithm works: the output is a sorted
permutation of the input.  In some sense it is an excellent test:
it will have considered all corner cases, including lists where all
numbers are the same, lists that are already sorted or reversely
sorted, etc.  If there is a bug in the sorting algorithm, most
likely it would be triggered and the model checker would produce a
scenario that would make it easy to find the source of the bug.
However, if the model checker does not find any bugs, we do not
know for sure that the algorithm works for lists with more than
five numbers or for lists that have values other than the numbers
1 through 5.  Still, we would expect that the likelihood that there
are bugs remaining in the sorting algorithm is small.

But model checking can also help with proving an algorithm correct.
The reason is that the correctness of many programs is built around
\emph{invariants}: predicates that must hold for every state in the
execution of a program.  A model checker can find violations of
those invariants when evaluating a model and provide valuable early
feedback to somebody who is trying to construct a proof, even an
informal one.
Throughout this book we will include examples
of such invariants as they often provide excellent insight into
why a particular algorithm works.

So what is CXL?
CXL is a concurrent programming language.  It was designed to teach
the basics of concurrent programming, but it is also useful for
testing new concurrent algorithms or even sequential and distributed
algorithms.  CXL programs are not intended to be ``run'' like programs
in most other programming languages---instead CXL programs are
model checked to test that the program has certain desirable
properties and does not suffer from bugs.

@@

CXL programs manipulate CXL values.
CXL values are recursively defined:
they include booleans (\texttt{False} and \texttt{True}),
integers (but not floating point numbers),
strings (enclosed by double quotes),
sets of CXL values,
and dictionaries that map CXL values to other CXL values.
%
Another type of CXL value is the \emph{atom}.  It is essentially
just a name.  An atom is denoted using a period followed by the
name.  For example, \texttt{.main} is an atom.

CXL makes extensive use of dictionaries.  Their syntax and properties
are a little different from Python.
A dictionary maps CXL values, known as \emph{keys}, to CXL values.
Unlike Python, any CXL value can be a key, including another
dictionary.
CXL dictionaries are written as
$\mathtt{dict}\{ k_0: v_0, ~ k_1: v_1, ~ ... \}$.
If \texttt{d} is a dictionary, and \texttt{k} is a key, then the
following expression retrieves the CXL value that $k$ maps to:
\begin{code}
\begin{verbatim}
d k
\end{verbatim}
\end{code}
This is unfamiliar to Python programers, but in CXL square brackets can be used
in the same way as parentheses, so you can express it in the more familiar form:
\begin{code}
\begin{verbatim}
d[k]
\end{verbatim}
\end{code}
However, if \texttt{d = dict\{ .count: 3 \}}, then you can write
can write \texttt{d.count} (which has value 3) instead of having to write
\texttt{d[.count]} (although both will work).
The meaning of \texttt{d a b c ...} is \texttt{((((d a) b) c) ...)}.

Tuples are special forms of dictionaries where the keys are
the indexes into the tuple.  For example, the tuple
\texttt{(5, False)} is the same CXL value as
\texttt{dict\{ 0: 5, 1: False \}}.
The empty tuple \texttt{()} is the same value as \texttt{dict\{\}}.
Note that this is different from the empty set, which is \texttt{\{\}}.
As in Python, you can create singleton tuples by including a comma.
For example, \texttt{(1,)}.

@@

The state of a process is called a \emph{context}: it contains the value of
its name tag, program counter, stack, and register.
The state of the CXL machine
consists of the value of its memory and the multiset (or \emph{bag}) of
contexts.  It is a multiset of contexts because two different processes can
be in the same state.
The initial state of the CXL memory is the empty dictionary, \texttt{()}.
The context bag has an initial context in it with name tag
\texttt{\_\_main\_\_/()}, pc 0, register \texttt{()}, and an empty stack.
Each machine instruction updates the state in some way.
Figure~\ref{fig:incstate} shows an example of a reachable state for 
the program in Figure~\ref{fig:inc}.

It may seem strange that there is only one memory location and that each
process has only one register.  However, this is not a limitation because
CXL values are unbounded trees.
Both the memory and the register of a process always contain
a dictionary that maps atoms to CXL values.  We call this a \emph{directory}.
A directory represents the state of a collection of variables named by the atoms.
%
Because directories are CXL values themselves,
directories can be organized into a tree.
Each node in the tree is then identified
by a sequence of atoms, like a path name in the file system hierarchy.  We call
such a sequence the \emph{address} of a CXL value, and it is relative to the
``root'' of the CXL value.

Compiling the code in Figure~\ref{fig:inc} results in the CXL machine code
listed in Figure~\ref{fig:inccode}.
The CXL machine is predominantly a \emph{stack machine}.
All instructions are atomically executed.
Most instructions pop values from the stack or push values onto the stack.
Initially there is one process (with name tag \texttt{\_\_init\_\_/()}
that starts executing at instruction 0 and keeps executing until the last
execution in the program to initialize the state.
In this case, the first instruction is a \texttt{JUMP} instruction that sets the
program counter to 39.
At program counter 1 is the code for the \texttt{incrementer} method.
All methods start with a \texttt{Frame} instruction and end with a \texttt{Return}
instruction.

@@

\begin{enumerate} \setcounter{enumi}{1}
\item the \texttt{Push} instruction pushes the constant 1
onto the stack of the process.
\item the \texttt{PushAddress} instruction pushes the address of
the shared variable \texttt{count} onto the stack.
\item The \texttt{Load} instruction pops the address of the
\texttt{count} variable and then pushes the value of the
\texttt{count} variable onto the stack.
\item \texttt{2-ary} is a \texttt{+} operation with 2 arguments.
It pops two values from the stack (1 and the value of \texttt{count}),
adds them and pushes the result back onto the stack.
\item The \texttt{PushAddress} instruction pushes the address of the \texttt{count}
variable onto the stack.
\item The \texttt{Store} instruction pops the address of a variable and pops
a CXL value (the sum of the \texttt{count} variable and 1) and updates the
\texttt{count} variable.
\end{enumerate}

Once initialization completes, any processes that were spawned can run.
You can think of CXL as trying every possible interleaving of processes executing
instructions.
Figure~\ref{fig:incoutput} shows the output produced by running CXL on the
\texttt{Up.cxl} program.
It starts by reporting that the assertion on line 9 in column 5 failed and that
\texttt{count} has value 1 instead of 0:
\texttt{Assertion failed (’assert’, ’Up.cxl’, 9, 5) 1}.
Next it reports an execution that failed this assertion.  The output has
four columns:
\begin{enumerate}
\item The name tag of the process;
\item The sequence of program counters of the CXL machine instructions that the process executed;
\item The current program counter of the process;
\item The contents of the shared memory.
\end{enumerate}

If we look at the middle three rows, we see that:
\begin{enumerate}
\item Process 0 executed instructions 1 through 7, loading the value of
\texttt{count} but stopping just before storing 1 into \texttt{count};
\item Process 1 executed instructions 1 through 12, storing 1 into
into \texttt{count} and storing \texttt{True} into \texttt{done[1]};
\item Process 0 continues execution, storing value 1 into \texttt{count}
and storing \texttt{True} into \texttt{done[0]}.
\end{enumerate}

@@

Hopefully you have started thinking of how to solve the concurrency
problem and you may already have prototyped some solutions.
In this chapter we will go through a few reasonable but broken attempts.
At the heart of the problem is that we would like make sure that, when
the \texttt{count} variable is being updated, no other process is
trying to do the same thing.  We call this a \emph{critical section}: a
set of instructions where only one process is allowed to execute at a
time.

Critical sections are useful when accessing a shared data
structure, particularly when that access requires multiple underlying
machine instructions.  A counter is a very simple example of
a data structure.  A more involved one would be access to a binary tree.
Adding a node to a binary tree, or re-balancing a tree, often requires
multiple operations.  Maintaining ``consistency'' is certainly much easier
(although not necessarily impossible) if during this time no other
process also tries to access the binary tree.

A critical section is often modeled as processes in an infinite loop
entering and exiting the critical section.
Figure~\ref{fig:csbarebones} shows the CXL code.
Here \texttt{@ncs} and \texttt{@cs} are \emph{labels}, each identifying
a location in the CXL machine code.  The first thing we need to
ensure is that there cannot be two processes at the critical section.
We would like to place an assertion at the \texttt{@cs} label that
specifies that only the current process can be there.  CXL in fact
supports this.  It has an operator \texttt{atLabel $L$}, where $L$
is the atom of the label (in this case, \texttt{.cs}).  The
operator returns a bag of name tags of processes executing at that
label.  The bag is represented by a dictionary that maps each element
in the bag to the number of times the element appears in the bag.
Method \texttt{atLabel} only exists for specification purposes---do not
use it in normal code.

The assertion also makes use of the \texttt{nametag()} operator that
returns the name tag of the current process.
If you run the code through CXL, the assertion should fail because
there is no code yet for entering and exiting the critical section.
